---
title: "SeaFlow Curation Report"
output: 
  html_document:
    toc: true
params:
  db: null
  curation_config_path: null
  save_path: null
  fig_width: 8
  fig_height: 4
  warning: false
  message: false
  error: false
  cache: false
  printcode: false
---

```{r, vscode-pre-setup, echo=FALSE, eval=FALSE}
# Run this chunk manually to set custom params in vscode

params_yaml <- "params:
  db: null
  curation_config_path: null
  save_path: null
  fig_width: 8
  fig_height: 4
  warning: false
  message: false
  error: false
  cache: false
  printcode: false"
params <- yaml::yaml.load(params_yaml)$params
```

```{r, setup, include=FALSE}
library(tidyverse)

knitr::opts_chunk$set(echo = params$printcode)
knitr::opts_chunk$set(fig.width = params$fig_width)
knitr::opts_chunk$set(fig.height = params$fig_height)
knitr::opts_chunk$set(warning = params$warning)
knitr::opts_chunk$set(message = params$message)

db <- params$db
curation_config_path <- params$curation_config_path
save_path <- params$save_path

# Don't run any chunks if we're missing required parameters
knitr::opts_chunk$set(eval = !(is.null(db) || is.null(curation_config_path) || is.null(save_path)))

# Plotting setup
col.t <- "deepskyblue3"
col.l <- "red3"

# Capture report date
report_date <- Sys.time()
attr(report_date, "tzone") <- "UTC"
```

## Cruise

```{r, cruise, results = "asis"}
cruise <- popcycle::get_cruise(db)
cat(cruise)
```

## Creation Date

```{r, creation-date, results = "asis"}
cat(lubridate::format_ISO8601(report_date, usetz = TRUE))
```

## Input Files

* popcycle database file = `r db`
* curation configuration file = `r curation_config_path`

```{r, load-data}
# Load curation configuration
curate <- readr::read_tsv(curation_config_path)

# load stat table
stat <- popcycle::get_stat_table(db)
stat$time <- lubridate::ymd_hms(stat$time)
stat$flag <- 0 # add a 'flag' column to the table or set to 0

# get bead fsc coordinates used in OPP filtration
filter_table <- popcycle::get_filter_table(db)
filter_plan_table <- popcycle::get_filter_plan_table(db)

filter_table <- inner_join(filter_table, filter_plan_table, by = join_by(id == filter_id))
filter_table <- filter_table %>%
  select(id, quantile, beads_fsc_small, start_date) %>%
  arrange(start_date, quantile)
filter_table <- popcycle::transformData(filter_table, column = "beads_fsc_small")
filter_dates <- filter_table %>%
  group_by(id) %>%
  summarise(start_date = unique(start_date)) %>%
  arrange(start_date)
filter_dates[nrow(filter_dates), "end_date"] <- max(stat$time)
filter_dates$end_date <- lead(filter_dates$start_date)
filter_dates[nrow(filter_dates), "end_date"] <- max(stat$time)
filter_table <- inner_join(filter_table, filter_dates[, c("id", "end_date")], by = "id")

fid <- unique(popcycle::get_filter_table(db)$id)[1]
fsc.beads <- popcycle::transformData(popcycle::get_filter_params_by_id(db, fid), column="beads.fsc.small")$beads.fsc.small

# load OPP table
opp <- popcycle::get_opp_table(db, outlier_join = FALSE)
opp <- as.data.frame(opp)
opp$flag <- 0 # add 'flag' column to the table
opp$event_rate <- opp[, "all_count"] / 180 # calculate event rate per sec (using total number of event recorded in 3 minutes)
```

## Outlier identification

### Instrument, Flag 1

#### Calibration and Testing Periods

```{r, instrument-calibration}
## Remove calibration or testing data ##
if (!is.na(curate$tcut1)) { # applies to only a few cruises
  df <- stat %>%
    filter(flag == 0) %>%
    group_by(time, file_id, quantile) %>%
    summarise(n_count = sum(n_count)) %>%
    ungroup() %>%
    as.data.frame()
  para <- "n_count"
  df$quantile <- as.factor(df$quantile)

  tcut1 <- as.POSIXct(curate$tcut1, format = "%Y-%m-%d %H:%M", tz = "UTC")

  if (curate$tcut1_gt_lt == "leq") {
    out1 <- which(df$time <= tcut1)
  } else if (curate$tcut1_gt_lt == "geq") {
    out1 <- which(df$time >= tcut1)
  }

  p <- df %>% ggplot() + geom_point(aes(time, .data[[para]], fill = quantile), pch=21, size=3, alpha=0.25) +
    geom_point(data=df[out1, ], aes(time, .data[[para]]), pch=21, size=3, alpha=1, fill=col.l)
  nout <- length(out1)

  if (!is.na(curate$tcut2)) { # one cruise has two time cut-offs
    tcut2 <- as.POSIXct(curate$tcut2, format = "%Y-&m-&d %H:%M", tz = "UTC")
    if (curate$tcut2_gt_lt == "leq") {
      out2 <- which(df$time <= tcut2)
    } else if (curate$tcut2_gt_lt == "geq") {
      out2 <- which(df$time >= tcut2)
    }

    p <- p +
      geom_point(data=df[out2, ], aes(time, .data[[para]]), pch=21, size=3, alpha=1, fill=col.l)

    nout <- length(out1) + length(out2)
  } # end second time cut

  print(p)

  print(paste0(round(100 * nout / nrow(df), 2), "% outliers"))

  # Set instrument issues flag (1) for time cutoffs
  id0.1 <- which(!is.na(match(stat[, "file_id"], unique(df[out1, "file_id"]))))
  stat[id0.1, "flag"] <- 1

  id0.2 <- which(!is.na(match(opp[, "file_id"], unique(df[out1, "file_id"]))))
  opp[id0.2, "flag"] <- 1

  if (!is.na(curate$tcut2)) {
    id0.12 <- which(!is.na(match(stat[, "file_id"], unique(df[out2, "file_id"]))))
    stat[id0.12, "flag"] <- 1

    id0.22 <- which(!is.na(match(opp[, "file_id"], unique(df[out2, "file_id"]))))
    opp[id0.22, "flag"] <- 1
  }
} else {
  cat("No calibration or testing periods in this cruise")
}
```

#### Flow Rate

```{r, instrument-flow-rate}
## Remove FLOW_RATE outliers
para <- "stream_pressure"
df <- stat %>%
  filter(flag == 0) %>%
  group_by(time) %>%
  filter(row_number() == 1) %>%
  as.data.frame()
out <- which(df[, para] > curate$lim1_2 | df[, para] < curate$lim1_1)

p <- df %>% ggplot() + geom_point(aes(time, .data[[para]]), pch=21, size=3, alpha=0.25, fill = "grey") +
  geom_point(data=df[out, ], aes(time, .data[[para]]), pch=21, size=3, alpha=0.25, fill = col.t) +
  geom_hline(yintercept = c(curate$lim1_1, curate$lim1_2), color=col.t, linetype = "dashed")
print(p)

print(paste0(round(100 * length(out) / nrow(df), 2), "% outliers"))

# Set instrument issues flag (1) for flow rate
id1.1 <- which(!is.na(match(stat[, "file_id"], unique(df[out, "file_id"]))))
stat[id1.1, "flag"] <- 1
id1.2 <- which(!is.na(match(opp[, "file_id"], unique(df[out, "file_id"]))))
opp[id1.2, "flag"] <- 1
```

#### Event Rate

```{r, instrument-event-rate}
## Remove EVENT_RATE outliers ##
para <- "event_rate"
df1 <- subset(opp, flag == 0)

out1 <- which(df1[, para] > curate$lim2_2 | df1[, para] < curate$lim2_1)

# low pass filter
if (length(out1) != 0) {
  df2 <- df1[-out1, ]
} else {
  df2 <- df1
}
out2 <- c()
if (nrow(df2) > 0) {
  model <- smooth.spline(df2$date, df2[, para])
  res <- residuals(model)
  out2 <- which(res < -curate$fact.sd2*sd(res) | res > curate$fact.sd2*sd(res))

  p <- df1 %>% ggplot() + geom_point(aes(date, .data[[para]]), pch=21, size=3, alpha=0.25, fill="grey") +
    geom_point(data=df1[out1, ], aes(date, .data[[para]]), pch=21, size=3, alpha=0.25, fill=col.t) +
    geom_hline(yintercept = c(curate$lim2_1, curate$lim2_2), color=col.t, linetype="dashed") +
    geom_point(data=df2[out2, ], aes(date, event_rate), pch=21, size=3, alpha=0.25, fill=col.l)
  print(p)
}

print(paste0(round(100 * length(c(out1, out2)) / nrow(df), 2), "% outliers"))

# Set instrument issues flag (1) for event rate
id2.1 <- which(!is.na(match(stat[, "file_id"], unique(c(df1[out1, "file_id"], df2[out2, "file_id"])))))
stat[id2.1, "flag"] <- 1
id2.2 <- which(!is.na(match(opp[, "file_id"], unique(c(df1[out1, "file_id"], df2[out2, "file_id"])))))
opp[id2.2, "flag"] <- 1
```

#### File Writing Errors

```{r, instrument-file-writing}
## Remove file writing issues ##
para <- "evt_count"
df1 <- subset(opp, flag == 0)

out1 <- which(df1[, para] > curate$lim3_2 | df1[, para] < curate$lim3_1)

# low pass filter
if (length(out1) != 0) {
  df2 <- df1[-out1, ]
} else {
  df2 <- df1
}
out2 <- c()

if (nrow(df2) > 0) {
  model <- smooth.spline(df2$date, df2[, para])
  res <- residuals(model)
  out2 <- which(res < -curate$fact.sd3*sd(res) | res > curate$fact.sd3*sd(res))

  p <- df1 %>% ggplot() + geom_point(aes(date, .data[[para]]), pch=21, size=3, alpha=0.25, fill="grey") +
    geom_point(data=df1[out1, ], aes(date, .data[[para]]), pch=21, size=3, alpha=0.25, fill=col.t) +
    geom_hline(yintercept = c(curate$lim3_1, curate$lim3_2), color=col.t, linetype="dashed") +
    geom_point(data=df2[out2, ], aes(date, .data[[para]]), pch=21, size=3, alpha=0.25, fill=col.l)
  print(p)
}

print(paste0(round(100 * length(c(out1, out2)) / nrow(df), 2), "% outliers"))

# Set instrument issues flag (1) for file writing
id3.1 <- which(!is.na(match(stat[, "file_id"], unique(c(df1[out1, "file_id"], df2[out2, "file_id"])))))
stat[id3.1, "flag"] <- 1
id3.2 <- which(!is.na(match(opp[, "file_id"], unique(c(df1[out1, "file_id"], df2[out2, "file_id"])))))
opp[id3.2, "flag"] <- 1
```

### Virtual Core, Flag 2

#### Beads

```{r, virtual-core-beads, fig.width=8, fig.height=8}
## Beads ##
phyto <- "beads"
if (!any(unique(stat$pop) == phyto)) {
  print(paste(phyto, "not found"))
} else {
  # threshold
  para <- "fsc_med"
  df1 <- subset(stat, flag == 0 & pop == phyto)
  out1 <- c()
  if (nrow(df1) > 0) {
    out1 <- which(df1[, para] > curate$lim4_2 | df1[, para] < curate$lim4_1)
    df1$log_abundance <- log10(df1$abundance)

    p <- df1 %>% ggplot() + geom_point(aes(time, .data[[para]], fill = log_abundance), pch=21, size=3, alpha=0.25) +
      geom_point(data=df1[out1, ], aes(time, .data[[para]]), pch=21, size=3, fill=col.t) +
      geom_hline(yintercept = c(curate$lim4_1, curate$lim4_2), color=col.t, linetype="dashed") +
      geom_segment(
        aes(x = start_date, xend = end_date, y = beads_fsc_small, yend = beads_fsc_small),
        color = "red",
        data = filter_table,
      ) +
      ggtitle(paste(phyto)) +
      scale_y_continuous(trans="log10") +
      scale_fill_viridis_c() +
      facet_grid(quantile ~ .)
    print(p)

    # Set virtual core issues flag (2) for beads
    id4.1 <- which(!is.na(match(stat[, "file_id"], unique(df1[out1, "file_id"]))))
    stat[id4.1, "flag"] <- 2
    id4.2 <- which(!is.na(match(opp[, "file_id"], unique(df1[out1, "file_id"]))))
    opp[id4.2, "flag"] <- 2
  }

  print(paste0(round(100 * length(out1) / nrow(df), 2), "% outliers"))
}
```

#### OPP/EVT Ratio

```{r, virtual-core-opp-evt-ratio}
## Remove OPP FILTRATION outliers ##
para <- "opp_evt_ratio"
df1 <- subset(opp, flag == 0)
out1 <- which(df1[, para] > curate$lim5_2 | df1[, para] < curate$lim5_1)
df1$quantile <- as.factor(df1$quantile)

# low pass filter
if (length(out1)!= 0) {
  df2 <- df1[-out1, ]
} else {
  df2 <- df1
}

out2 <- c()

if (nrow(df2) > 0) {

  model <- smooth.spline(df2$date, df2[, para])
  res <- residuals(model)
  out2 <- which(res < -curate$fact.sd5*sd(res) | res > curate$fact.sd5*sd(res))

  p <- df1 %>% ggplot() + geom_point(aes(date, .data[[para]], fill = quantile), pch=21, size=3, alpha=0.25) +
    geom_point(data=df1[out1, ], aes(date, .data[[para]]), pch=21, size=3, fill=col.t) +
    geom_hline(yintercept = c(curate$lim5_1, curate$lim5_2), color=col.t, linetype="dashed") +
    geom_point(data=df2[out2, ], aes(date, .data[[para]]), pch=21, size=3, fill=col.l) +
    scale_y_continuous(trans="log10")
  print(p)
}

print(paste0(round(100 * length(c(out1, out2)) / nrow(df), 2), "% outliers"))

# Set virtual core issues flag (2) for opp_evt_ratio
id5.1 <- which(!is.na(match(stat[, "file_id"], unique(c(df1[out1, "file_id"], df2[out2, "file_id"])))))
stat[id5.1, "flag"] <- 2
id5.2 <- which(!is.na(match(opp[, "file_id"], unique(c(df1[out1, "file_id"], df2[out2, "file_id"])))))
opp[id5.2, "flag"] <- 2
```

### Classification, Flag 3

#### Size, Prochlorococcus

```{r, classification-prochlorococcus}
### Remove Prochlorococcus size outliers
if (!is.na(curate$lim6_1)) {   # Some cruises don't gate for Pro
  phyto <- "prochloro"
  if (!any(unique(stat$pop) == phyto)) {
    print(paste(phyto, "not found"))
  }

  # threshold
  para <- "diam_mid_med"
  df1 <- subset(stat, flag == 0 & pop == phyto)
  out1 <- which(df1[, para] > curate$lim6_2 | df1[, para] < curate$lim6_1)
  df1$quantile <- as.factor(df1$quantile)

  # low pass filter
  if (length(out1)!=0) {
    df2 <- df1[-out1, ]
  } else {
    df2 <- df1
  }

  out2 <- c()

  if (nrow(df2) > 0) {
    model <- smooth.spline(df2$time, df2[, para])
    res <- residuals(model)
    out2 <- which(res < -curate$fact.sd6*sd(res) | res > curate$fact.sd6*sd(res))

    p <- df1 %>% ggplot() + geom_point(aes(time, .data[[para]], fill = quantile), pch=21, size=3, alpha=0.25) +
      geom_point(data=df1[out1, ], aes(time, .data[[para]]), pch=21, size=3, fill=col.t) +
      geom_hline(yintercept = c(curate$lim6_1, curate$lim6_2), color=col.t, linetype="dashed") +
      geom_point(data=df2[out2, ], aes(time, .data[[para]]), pch=21, size=3, fill=col.l) +
      ggtitle(paste(phyto))
    print(p)
  }

  print(paste0(round(100 * length(c(out1, out2)) / (nrow(df1)), 2), "% outliers"))

  # Set classification issues flag (3) for pro
  id6 <- which(!is.na(match(stat[, "file_id"], unique(c(df1[out1, "file_id"], df2[out2, "file_id"])))))
  stat[id6, "flag"] <- 3
} # End Pro conditional
```

#### Size, Synechococcus

```{r, classification-synechococcus}
### Remove Synechococcus size outliers
phyto <- "synecho"
if (!any(unique(stat$pop) == phyto)) {
  print(paste(phyto, "not found"))
}

# threshold
para <- "diam_mid_med"
df1 <- subset(stat, flag == 0 & pop == phyto)
out1 <- which(df1[, para] > curate$lim7_2 | df1[, para] < curate$lim7_1)
df1$quantile <- as.factor(df1$quantile)

# low pass filter
if (length(out1)!= 0) {
  df2 <- df1[-out1, ]
} else {
  df2 <- df1
}

out2 <- c()

if (nrow(df2) > 0) {
  model <- smooth.spline(df2$time, df2[, para])
  res <- residuals(model)
  out2 <- which(res < -curate$fact.sd7*sd(res) | res > curate$fact.sd7*sd(res))

  p <- df1 %>% ggplot() + geom_point(aes(time, .data[[para]], fill = quantile), pch=21, size=3, alpha=0.25) +
    geom_point(data=df1[out1, ], aes(time, .data[[para]]), pch=21, size=3, fill=col.t) +
    geom_hline(yintercept = c(curate$lim7_1, curate$lim7_2), color=col.t, linetype="dashed") +
    geom_point(data=df2[out2, ], aes(time, .data[[para]]), pch=21, size=3, fill=col.l) +
    ggtitle(paste(phyto))
  print(p)
}

print(paste0(round(100 * length(c(out1, out2)) / nrow(df1), 2), "% outliers"))

# Set classification issues flag (3) for syn
id7 <- which(!is.na(match(stat[, "file_id"], unique(c(df1[out1, "file_id"], df2[out2, "file_id"])))))
stat[id7, "flag"] <- 3
```

#### Abundance, All Populations

```{r, classification-abundance}
## Abundance:  all populations ##
if (curate$spar.8 == "missing" || curate$spar.8 == "NULL") {
  curate$spar.8 <- NULL
}

df <- subset(stat, flag == 0 & quantile == 2.5)   # Easier to catch outliers when you don't consider all 3 quantiles
para <- "abundance"
all <- unique(df$pop)
phyto <- all[-which(all == "unknown" | all == "beads")]

OUT <- NULL
for (i in c(phyto)) {
  df1 <- subset(df, pop == i)
  df1$quantile <- as.factor(df1$quantile)
  if (nrow(df1) > 0) {
    model <- smooth.spline(df1$time, df1[, para], spar = curate$spar.8)
    res <- residuals(model)
    out1 <- which(res < -curate$fact.sd8*sd(res) | res > curate$fact.sd8*sd(res))
    out <-  as.vector(unlist(data.frame(filename=unique(df1[out1, "file_id"]))))
    OUT <- unique(c(OUT, out))

    p <- df1 %>% ggplot() + geom_point(aes(time, .data[[para]]), fill = "grey", pch=21, size=3, alpha=0.25) +
      geom_point(data=df1[out1, ], aes(time, .data[[para]]), pch=21, size=3, alpha=1, fill=col.l) +
      ggtitle(paste(i))
    print(p)

    print(paste0(round(100 * length(out) / nrow(df1), 2), "% outliers"))
  }
}

# Set classification issues flag (3) for adundance
id8 <- which(!is.na(match(stat[, "file_id"], OUT)))
stat[id8, "flag"] <- 3
```

## Curated data

### Abundance

```{r, curated-plot-abundance, fig.width=8, fig.height=8}
clean <- subset(stat, flag == 0 & pop != "beads" & pop != "unknown")
if (nrow(clean) > 0) {
  p <- popcycle::plot_time(clean, param = "abundance")
  print(p)
}
```

### Diameter (diam_mid_med)
```{r, curated-plot-size, fig.width=8, fig.height=8}
clean <- subset(stat, flag == 0 & pop != "beads" & pop != "unknown")
if (nrow(clean) > 0) {
  p <- popcycle::plot_time(clean, param = "diam_mid_med")
  print(p)
}
```

```{r, curated-total-outliers}
print(paste0(round(100 * nrow(stat[which(stat$flag != 0), ]) / nrow(stat), 2), "% outliers"))
```

```{r, write-table}
##########################
### Save outlier table ###
##########################
df <- stat[match(unique(stat$file_id), stat$file_id), ]
outliers <- data.frame(file = df$file_id, flag = df$flag)
readr::write_tsv(outliers, file = save_path)
```
